---
title: "Installing Geospatial Libraries"
author: "Albert Y. Kim"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installing Geospatial Libraries}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(SpatialEpi)
```

The next steps will install the open source geospatial libaries
necessary for your computer to read in ESRI shapefiles. Fortunately you
only have to do this once.

## Installation Instructions

### Mac Users

-   Install either Xcode (available for free in the App Store) or the
    [Command Line Tools for
    Xcode](https://developer.apple.com/download/all/). You'll need to have
    a (free) Apple ID.

-   In your Mac menu bar go to System Preferences... -\> Security &
    Privacy -\> unlock the lock on the bottom left of panel -\> select
    "Allow apps downloaded from": Anywhere -\> relock the lock

-   Open Terminal by going to a Finder window -\> Applications -\>
    Utilities -\> Terminal. You open new tabs in Terminal by pressing
    command+T.

-   In Terminal

    -   Run `sudo xcodebuild -license`, scroll to the bottom and type
        `agree` to agree to the licence.
    -   Open a new tab and run `xcode-select --install`

-   Install Macports as per the
    [Quickstart](https://www.macports.org/install.php) instructions.

    -   You have already completed steps 1 and 2.
    -   To know what version of OS X you have click on the Apple icon on
        the top left of the menu bar -\> About This Mac

-   Install the [`gdal`](https://gdal.org/) and
    [`geos`](https://trac.osgeo.org/geos/) geospatial libraries.

    -   In Terminal open a new tab and run `sudo port install gdal`
    -   In Terminal open a new tab and run `sudo port install geos`

-   Install the `rgdal` package to interface the `gdal` geospatial
    library with R

    -   Under
        [Downloads](https://CRAN.R-project.org/package=rgdal)
        select "Package Source". It should be named `rgdal_X.Y-Z.tar.gz`
        where `X,Y,Z` denote the version number.
    -   In Terminal open a new tab and run
        `R CMD INSTALL   ~/Downloads/rgdal_X.Y-Z.tar.gz` where `X,Y,Z`
        are changed to match the name of the file you just downloaded

-   Install the `rgeos` package to interface the `geos` geospatial
    library with R

    -   Under
        [Downloads](https://CRAN.R-project.org/package=rgeos)
        select "Package Source". It should be named `rgeos_X.Y-Z.tar.gz`
        where `X,Y,Z` denote the version number.
    -   In Terminal open a new tab and run
        `R CMD INSTALL   ~/Downloads/rgeos_X.Y-Z.tar.gz` where `X,Y,Z`
        are changed to match the name of the file you just downloaded

### Windows Users

-   Follow the instructions in [Quick Start for OSGeo4W
    Users](https://trac.osgeo.org/osgeo4w/)
-   In R, run
    `install.packages("rgeos", repos="https://R-Forge.R-project.org", type="source")`

### Linux Users

I'm going to assume you know what you're doing!

## Testing

After installing the prerequisite packages, test that both the `rgdal`
and `rgeos` packages are installed by running

```{r, eval=FALSE}
library(rgdal)
library(rgeos)
```
